---
title: "Prediction Assignment Writeup"
author: "Alexander Seifert"
output: html_document
---

```{r, echo=FALSE}
library(knitr); opts_chunk$set(cache=TRUE)
```

This project builds a machine learning algorithm to predict activity quality from activity monitors. By creating a [random forest](https://en.wikipedia.org/wiki/Random_forest) we achieve an accuracy of `0.991` (out of sample error `0.009`) on the validation set (60/40 split) and `100%` accuracy on the test set.


## Introduction

Human activity recognition research has traditionally focused on discriminating between different activities, i.e. to predict "which" activity was performed at a specific point in time (like with the Daily Living Activities dataset above).

In this project, we will use data from accelerometers on the belt, forearm, arm, and dumbell of 6 test participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways:

* exactly according to the specification (Class A)
* throwing the elbows to the front (Class B)
* lifting the dumbbell only halfway (Class C)
* lowering the dumbbell only halfway (Class D) and
* throwing the hips to the front (Class E)

(More information on the study that collected the data is available on the Groupware\@LES [Human Activity Recognition project website](http://groupware.les.inf.puc-rio.br/har#ixzz34oK3Rck9).)

We will try to predict from the sensory data the way in which the exercise was performed (i.e. class A to E).

## Data Preprocessing

First we set the seed, so our analysis is reproducible. We then load the training and testing data. Some of the data contained "#DIV/0!" entries, which most likely came from an illegal division by 0. We treat those as NAs.

```{r}
set.seed(42)
library(caret)

training.raw = read.csv("pml-training.csv", na.strings = c("", "NA", "#DIV/0!"))
testing.raw  = read.csv("pml-testing.csv", na.strings = c("", "NA", "#DIV/0!"))
```


As a next step, we remove near zero variance predictors.

```{r}
nzv  = nearZeroVar(training.raw, saveMetrics=TRUE)
omit = which(nzv$nzv==TRUE)
training.new = training.raw[,-omit]
```

Let's weed out the NA heavy colums
```{r}
NAs = apply(training.new, 2, function(x) { sum(is.na(x)) })
qplot(NAs, binwidth = max(NAs)/30)
good.cols = names(NAs[NAs == 0])
```


From the remaining variables, we keep only those coming from sensory input:
```{r}
predictors = grep("(_belt|_arm|_dumbbell|_forearm)", good.cols, value=T)
```


Now we create a new training set with only the predictors and the output (`classe`)
```{r}
training.new = cbind(training.new[,c(predictors, "classe")])
```

Having done these steps, we can move on to model creation.


## Model Creation

First, we create a 60/40 data split for training/validation set:
```{r}
intrain  = createDataPartition(training.new$classe, p = 0.6, list = FALSE)
training = training.new[intrain,]
validate = training.new[-intrain,]
```

Now we can train our model. We will use [4-fold cross validation](https://en.wikipedia.org/wiki/Cross-validation_(statistics)) and fit a random forest to the training set.
```{r}
trc = trainControl(method = "cv", number = 4, allowParallel=T)
rf.fit = train(classe ~ .,
               data = training,
               method = "rf",
               trControl = trc)

```


## Model Evaluation

Finally, we can use our validation set to evaluate our model along different dimensions:

```{r}
rf.pred = predict(rf.fit, validate)
confusionMatrix(rf.pred, validate$classe)
```

With the accuracy being `0.991` we get an out of sample error of `1 - 0.991 = 0.009`.


## Test Output

We predict the test answers
```{r}
testing = testing.raw[, names(training)[-53]]
answers = predict(rf.fit, testing)

answers
```


… and write those answers to disk:
```{r}
pml_write_files = function(x){
        n = length(x)
        for(i in 1:n){
                filename = paste0("problem_id_",i,".txt")
                write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
        }
}
pml_write_files(answers)
```
